---
title: "Tidy data analysis for demography using R"
author: Rob J Hyndman
date: 26 June 2024
titlegraphic: bg-06.png
toc: true
abstract: I will introduce the vital package which allows analysis of demographic data using tidy tools. The package uses a variation of tsibble objects as the main data class, so all of the infrastructure available for tsibble and tibble objects can also be used with vital objects. Data may include births, deaths, mortality, fertility, population and migration data. Functions for plotting, smoothing, modelling and forecasting data are included. Models include the classical Lee-Carter model as well as functional data models. Future plans include replicating all of the models available in the demography and StMoMo packages. The package is available at https://pkg.robjhyndman.com/vital/.
format:
  presentation-beamer:
    keep-tex: true
    template-partials:
        - before-title.tex
execute:
  cache: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: load-packages
#| echo: false
library(knitr)
library(ggplot2)
library(dplyr)
library(vital)
aus <- aus_mortality |>
  mutate(State = Code) |>
  as_vital(index = Year,
    key = c(Age, Sex, State),
    .age = "Age",
    .sex = "Sex",
    .deaths = "Deaths",
    .population = "Exposure"
  ) |>
  arrange(desc(State), Year, Age, Sex) |>
  select(-Code) |>
  filter(Sex != "total", !(State %in% c("ACT","ACTOT","NT")))

# Colours to be viridis for continuous scales and Okabe for discrete scales
options(
  ggplot2.continuous.colour="viridis",
  ggplot2.continuous.fill = "viridis",
  ggplot2.discrete.colour = c("#D55E00", "#0072B2","#009E73", "#CC79A7", "#E69F00", "#56B4E9", "#F0E442"),
  ggplot2.discrete.fill = c("#D55E00", "#0072B2","#009E73", "#CC79A7", "#E69F00", "#56B4E9", "#F0E442")
)
```


# Vital objects

## tibble objects
\placefig{8.3}{0}{width=2.5cm}{tibble}
\placefig{11.1}{0}{width=2.5cm}{tidyverse}

\alert{Australian Deaths 1901--2020}

\fontsize{10}{10.5}\sf

```{r}
#| label: tibble
tibble::as_tibble(aus) |>
  arrange(desc(State), Year, Age, Sex)
```

## tsibble objects
\placefig{8.3}{0}{width=2.5cm}{tsibble}
\alert{Australian Deaths 1901--2020}

\fontsize{10}{10.5}\sf

```{r}
#| label: tsibble
tsibble::as_tsibble(aus) |>
  arrange(desc(State), Year, Age, Sex)
```

\only<2-3>{\begin{textblock}{4}(11.5, 1)
\fontsize{11}{12}\sf
\begin{block}{Variables}
Index:
  \begin{itemize}
    \item \texttt{Year}
  \end{itemize}
Keys:
  \begin{itemize}
    \item \texttt{Age}
    \item \texttt{Sex}
    \item \texttt{State}
  \end{itemize}
\end{block}
\end{textblock}}

\only<3>{\begin{textblock}{4}(11.5, 7.1)
\fontsize{11}{12}\sf
\begin{alertblock}{}
Every row must have a unique combination of Index and Keys
\end{alertblock}
\end{textblock}
}

## vital objects
\placefig{8.3}{0}{width=2.5cm}{vital-hex}
\alert{Australian Deaths 1901--2020}

\fontsize{10}{10.5}\sf

```{r}
#| label: vital
aus
```

\begin{textblock}{4}(11.5, 1.)
\fontsize{11}{12}\sf
\begin{block}{Variables}
Index:
  \begin{itemize}
    \item \texttt{Year}
  \end{itemize}
Keys:
  \begin{itemize}
    \item \texttt{Age}
    \item \texttt{Sex}
    \item \texttt{State}
  \end{itemize}
\end{block}
\end{textblock}

\begin{textblock}{4}(11.5, 7.1)
\fontsize{11}{12}\sf
\begin{alertblock}{}
Every row must have a unique combination of Index and Keys
\end{alertblock}
\end{textblock}

\begin{textblock}{4}(11.5, 4.8)
\fontsize{11}{12}\sf
\begin{block}{}
Variables denoting age, sex, deaths, births and population can also be specified.
\end{block}
\end{textblock}

# Using the Human Mortality and Fertility Databases

## Human Mortality Database

\begin{textblock}{3.02}(11.98,-0.13)
\begin{alertblock}{}\hfill
\textbf{mortality.org}
\end{alertblock}
\end{textblock}
\fullwidth{hmd}

## Human Fertility Database

\begin{textblock}{4.2}(10.8,-0.13)
\begin{alertblock}{}\hfill
\textbf{humanfertility.org}
\end{alertblock}
\end{textblock}
\fullwidth{hfd}


## HMD imports
\fontsize{10}{10}\sf

```{r}
#| eval: false
norway <- read_hmd(
  country = "NOR",
  username = "Nora.Weigh@mymail.com",
  password = "FF!5xeEFa6"
)
norway_births <- read_hmd(
  country = "NOR",
  username = "Nora.Weigh@mymail.com",
  password = "FF!5xeEFa6",
  variables = "Births"
)
```

\fontsize{14}{15}\sf\vspace*{-0.2cm}

* Uses `HMDHFDplus` package to handle the downloads.
* Default variables: `Deaths`, `Exposures`, `Population`, `Mx`
* Only 1x1 data supported.
* `read_hmd_files()` and `read_hfd_files()` allow reading of downloaded files.


# Plots
## Rainbow plots
\fontsize{10}{11}\sf

```{r}
#| label: plot0
#| fig-height: 4.3
aus |> autoplot(Mortality) + scale_y_log10()
```

## Rainbow plots
\fontsize{10}{11}\sf

```{r}
#| label: plot1
#| fig-height: 4.3
aus |> filter(State == "VIC") |> autoplot(Mortality) + scale_y_log10()
```

## Rainbow plots
\fontsize{10}{11}\sf

```{r}
#| label: plot2
#| fig-height: 3.5
aus |> filter(State == "VIC") |>
  mutate(Exposure = if_else(Sex == "female", -Exposure, Exposure)) |>
  autoplot(Exposure) +
  facet_grid(State ~ Sex, scales = "free_x") + coord_flip()
```

# Life tables and life expectancy

## Life tables
\fontsize{10}{11}\sf

```{r}
#| label: life-tables
life_table(aus)
```

## Life expectancy

\fontsize{10}{11}\sf

```{r}
#| label: life-expectancy
life_expectancy(aus)
```


## Life expectancy

\fontsize{10}{11}\sf

```{r}
#| label: life-expectancy2
#| fig-height: 3.5
life_expectancy(aus) |>
  ggplot(aes(x = Year, y = ex, colour = State)) +
  geom_line(linewidth = 1) +
  facet_grid(. ~ Sex)
```

# Mortality models

## Benchmark models
\fontsize{10}{11}\sf

```{r}
#| label: benchmarks
fit <- aus |>
  model(
    naive = FNAIVE(Mortality),
    mean = FMEAN(Mortality)
  )
fit
```

## Benchmark models
\fontsize{10}{11}\sf

```{r}
#| label: benchmarks2
fit |>
  filter(Sex == "female", State == "NSW") |>
  select(mean) |>
  report()
```

## Benchmark models
\fontsize{10}{11}\sf

```{r}
#| label: benchmarks3
#| fig-height: 3.7
fit |>
  select(mean) |>
  autoplot() + scale_y_log10()
```

## Lee-Carter models
\fontsize{10}{11}\sf

```{r}
#| label: lc1
fit <- aus |>
  model(
    naive = FNAIVE(Mortality),
    mean = FMEAN(Mortality),
    lc = LC(log(Mortality))
  )
fit
```

## Lee-Carter models
\fontsize{10}{10}\sf

```{r}
#| label: lc2
fit |>
  filter(Sex == "female", State == "NSW") |>
  select(lc) |>
  report()
```

## Lee-Carter models
\fontsize{10}{9}\sf\vspace*{-0.2cm}

```
Age functions
# A tibble: 101 × 3
    Age    ax     bx
  <int> <dbl>  <dbl>
1     0 -4.07 0.0155
2     1 -6.20 0.0221
3     2 -6.89 0.0199
# i 98 more rows

Time coefficients
# A tsibble: 120 x 2 [1Y]
   Year    kt
  <int> <dbl>
1  1901 109.
2  1902 111.
3  1903 108.
# i 117 more rows

Time series model: RW w/ drift

Variance explained: 86.61%
```

## Lee-Carter models
\fontsize{10}{11}\sf

```{r}
#| label: lc3
#| fig-height: 3.7
fit |>
  filter(State == "NSW") |>
  select(lc) |>
  autoplot()
```

## Lee-Carter models
\fontsize{10}{11}\sf

```{r}
#| label: lc4
fit |> select(lc) |> age_components()
```


## Lee-Carter models
\fontsize{10}{11}\sf

```{r}
#| label: lc5
fit |> select(lc) |> time_components()
```


## Functional data models

## Coherent FDM (generalizing Lee-Li)

# Future plans
## Future plans

- Stochastic population forecasting (as per Booth-Hyndman 2008)
- StMoMo
- other tools in the demography package
- demography package to be maintained but not developed
